Soundness Proofs for Iterative Deepening

### 2015-11-10 pldi-16/proof/convergence.tex
finish convergence proof, i think

% TODO: Here's what to do.
% Induct on the PPs by the order of next(my_PP) in the trace.
% Inductive assumption gives that all prior other data race PPs with next(other_PP) before next(my_PP) are reachable,
% Is there a prior data race PP with next(other_PP) *AFTER* next(my_PP)?
% If not, then we are done, and this one is reachable.
% I

\newcommand\pbh{\ensuremath{p_{\beta{}h}}}
\newcommand\paj{\ensuremath{p_{\alpha{}j}}}
\newcommand\pbk{\ensuremath{p_{\beta{}k}}}
%\newcommand\tak{\ensuremath{t_{\alpha{}k}}}
%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
\newcommand\tbh{\ensuremath{t_{\beta{}h}}}
\newcommand\tbk{\ensuremath{t_{\beta{}k}}}
\newcommand\coalesce[1]{\ensuremath{\mathsf{coalesce}(#1)}}

@@

It would be straightforward to show that any other data races in $\ppnext\pai'$, which would be discovered in this interleaving,
could be reordered to after $\ppothers\pai'$,
eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
Unfortunately, $\mathcal{I}'$ is not reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
Hence, we must perform the same ``coalescing'' for each thread in $\ppothers\pai'$ that we did for $\alpha$,
which we will abbreviate:
\[
	\mathcal{I}'' = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
\]
For each other thread $\beta$, $\coalesce{\ppothers\pai'}$ will contain a single transition,
starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
and ending with the next sync API PP, which we'll call $\pbk$.
It follows from sync API PPs being reachable by definition,
and from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
that $\mathcal{I''}$ is reachable.

%%%%%%%%%%%%%%%%% this didn't work out
%We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
%Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
%i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions\footnote{
%Why use this strange ordering, rather than simply inducting by the order of preemptions?
%Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
%It is also true that the $\ppnext{\pai}$ order does not necessarily match the PPs' reachability dependence relation, but it is easier to cover the exception cases using this ordering rather than vice versa.}.
%
%\begin{figure}[t]
%	TODO - example of why ppnext
%	\caption{TODO}
%	\label{fig:why-ppnext}
%\end{figure}
%%%% well, the way we do it, might not be the first reachable one either...
%
%\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
%For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj} \prec \ppnext{\pai} \in \mathcal{I}$, $\pbj$ is reachable.
%To show $\pai$ is reachable, the proof has two parts.
%First, we must handle any preceding data-race PPs $\pgk$ with $\ppnext{\pai} \prec \ppnext{\pgk}$.
%Second, we must show that when {\pai} is removed, a data-race including $\ppinstr{\pai}$ will be identified.
%%%%%%%%%%%%%%%%%%%

% TODO

\begin{itemize}
	\item
	\item
\end{itemize}
\end{proof}

>>>>

\newcommand\pbh{\ensuremath{p_{\beta{}h}}}
\newcommand\paj{\ensuremath{p_{\alpha{}j}}}
\newcommand\pbk{\ensuremath{p_{\beta{}k}}}
\newcommand\tbh{\ensuremath{t_{\beta{}h}}}
\newcommand\tgk{\ensuremath{t_{\gamma{}k}}}
\newcommand\tgka{\ensuremath{t_{\gamma{}k1}}}
\newcommand\tgkb{\ensuremath{t_{\gamma{}k2}}}
\newcommand\tdl{\ensuremath{t_{\delta{}l}}}
\newcommand\coalesce[1]{\ensuremath{\mathsf{coalesce}(#1)}}

@@

It would be straightforward to show that any other data races in $\ppnext\pai'$, which would be discovered in this interleaving,
could be reordered to after $\ppothers\pai'$,
eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
Unfortunately, $\mathcal{I}'$ is not necessarily reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
Hence, we must perform the same ``coalescing'' for each thread in $\ppothers\pai'$ that we did for $\alpha$,
which we will abbreviate:
\[
	\mathcal{I}'' = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
\]
For each other thread $\beta$, $\coalesce{\ppothers\pai'}$ will contain a single transition,
starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
and ending with the next sync API PP in $\beta$, which we'll call $\pbk$.
It follows from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
and from sync API PPs being reachable by definition,
that $\mathcal{I''}$ is reachable.

Again, the only certainly-executed instruction by any thread after $\tai$ is $\ppinstr\pai$.
It is even possible that neither $\ppinstr\pai$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
until a different data race PP between two other threads is discovered,
as shown in Figure~\ref{fig:threethreads}.

\begin{figure}[t]
	TODO % TODO
	\caption{In this example, even though thread 1's data-race PP must occur first in a buggy interleaving, it cannot be reached until after Iterative Deepening first reaches the later-occurring data-race PP between threads 2 and 3.}
	\label{fig:threethreads}
\end{figure}

Now, we show by induction that a data race on $\ppinstr\pai$ will be reachable.
We assume that a state space $\mathcal{S}$ containing $\mathcal{I}''$,
plus $n$ unique data race PPs among $\ppnext\pai'$ and $\ppothers\pai$,
will be reachable.
(In the base case, the number of new data-race PPs is 0, and $\mathcal{I}''$'s reachability is justified above.)
We must show that in this state space, if the $\ppinstr\pai$ data race is not reachable,
a new unique data-race PP nevertheless will be.

By the definition of fully-relevant interleaving, it must be possible for some other thread $\beta$ to execute an instruction that conflicts with $\ppinstr\pai$.
%
If this conflict is not observed in $\mathcal{S}$,
it means that some preceding thread communication (involving any two threads $\gamma$ and $\delta$, either potentially the same or different as $\alpha$ or $\beta$) must exist to have suppressed it,
and that not all possible interleavings of that communication were tested in $\mathcal{S}$.
%
By the converse of our definition of DPOR, if a possible (conflicting, concurrent) thread interleaving is not tested,
% XXX: Handwave? I'm not entirely convinced by this step. How does it justify skipping all the way to the first possible "reachable" data race among the chain of PPs needed to expose \beta's communication with \ppinstr\pai?
then there must exist some transition with a necessary site of interleaving in the middle of it, rather than at its boundaries.

Let $\tgk$ be this transition, $\tgka$ and $\tgkb$ be the result of splitting at that site, and $\tdl$ be the communicating transition of the other thread.
%
If the interleaving $\{\tgka, \tdl', \tgkb'\}$ can produce a different program behaviour than $\{\tgka, \tgkb, \tdl\}$,
then again by the soundness of DPOR, $\tgkb$ and $\tdl$ must contain some shared memory conflict.
By the maximal state space assumption, all sync API PPs are enabled, so the locks held by $\gamma$ cannot change during $\tgk$,
and if $\tdl$ can be reordered to the middle of $\tgk$, the locksets of $\gamma$ and $\delta$ cannot overlap,
and likewise there is no sync-enforced MHB relation.
Hence the memory conflict between $\tdl$ and $\tgkb$ will be identified as a data-race.
Because $\tgk$ was split such that the first instruction of $\tgkb$ was not previously a PP, the data-race was not already enabled as a PP.

Hence either a data-race including $\ppinstr\pai$ will be identified,
or an infinite/nonterminating sequence of other unique data-race PPs will be identified,
but by the halting assumption, this would constitute an infinite loop bug, which is sufficient.
(Alternatively, for any program with a finitely-sized instruction listing, the number of unique instruction pairs is finite.)
This finishes our inner induction.
Hence $\pai$ is reachable, finishing our outer induction.
Hence all PPs in $\mathcal{I}$ are reachable.
\end{proof}

%%%%%%%%%%%%%%%%% this didn't work out
%We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
%Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
%i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions\footnote{
%Why use this strange ordering, rather than simply inducting by the order of preemptions?
%Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
%It is also true that the $\ppnext{\pai}$ order does not necessarily match the PPs' reachability dependence relation, but it is easier to cover the exception cases using this ordering rather than vice versa.}.
%
%\begin{figure}[t]
%	\caption{to do}
%	\label{fig:why-ppnext}
%\end{figure}
%%%% well, the way we do it, might not be the first reachable one either...
%
%\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
%For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj} \prec \ppnext{\pai} \in \mathcal{I}$, $\pbj$ is reachable.
%To show $\pai$ is reachable, the proof has two parts.
%First, we must handle any preceding data-race PPs $\pgk$ with $\ppnext{\pai} \prec \ppnext{\pgk}$.
%Second, we must show that when {\pai} is removed, a data-race including $\ppinstr{\pai}$ will be identified.
%%%%%%%%%%%%%%%%%%%
